\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt] \huge Deep learning transcriptomic model for prediction of pan-drug chemotherapeutic sensitivity \\

### 2020-06-30 manuscript/manuscript.tex
updated figure caption
discussion needs to be fixed

% Useful packages
\usepackage{amsmath, graphicx, natbib, multicol, booktabs, float, abstract, placeins}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\bibliographystyle{unsrt}

@@

\begin{document}
% Abstract
\twocolumn[
	\maketitle
	\begin{abstract}
		Precision oncology approaches have recently begun to inform clinical decision-making in the context of chemotherapeutic regimen. High-throughput technology has enabled machine learning algorithms to utilize molecular characteristics of tumours to generate personalized treatments, revealing relationships between genomic alterations and drug sensitivity. However, precision oncology studies have yet to generate a predictive biomarker incorporating gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. Here we show that a 10 hidden layer neural network accurately classifies pan-cancer cell lines into responder and non-responder groups with respect to a pan-drug dataset with 90.8\% accuracy (AUC = 0.872 $\pm$ 0.054). From unsupervised clustering algorithms, we found that a cohort of cell line gene expression data from the Genomics of Drug Sensitivity in Cancer could be clustered into two response groups with significant difference in pan-drug chemotherapeutic sensitivity. Due to the high dimensional nature of the microarray data, we used the Boruta feature selection algorithm was used to identify genes of the highest predictive value; the 300 genes selected was enriched primarily for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using the selected genes, a deep learning neural network was developed to predict response groups. The high accuracy of the neural network validates our postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it suggests the potential utility of similar combinatorial biomarkers for the selection of potent candidate drugs. Given that not all tumours present with targetable features and are treated with less targeted and consequently more cytotoxic chemotherapies, our computational framework allows for the identification of candidate drugs for treatments that maximize therapeutic response and minimize cytotoxic burden. \vspace{1em}

@@

While previous studies have developed predictive biomarkers for specific drugs, the utility of these biomarkers is limited to particular patients and clinical contexts \cite{drug_sense}. That is, these studies are limited in terms of clinical generalizability to different chemotherapeutic regimens. However, a pan-drug predictive biomarker may provide significant clinical utility in the selection of candidate therapies for particular patients. Such a biomarker could be developed if tumours with similar expression have similar drug responses across all chemotherapies, with few exceptions. The availability of pan-cancer cell line databases with \textit{in vitro} drug sensitivity analyses along with accompanying gene expression profiling provides an ideal model for such analyses \cite{gdsc}. However, the majority of previous drug sensitivity predictive biomarkers built on cell line databases have utilized classical machine learning combinatorial techniques, which fail to capture the dimensionality of available transcriptomic data. Furthermore, advanced deep learning algorithmic approaches that are capable of handling such dimensionality often fail to allow interpretability, and consequently require transcriptomic data that is clinically infeasible \cite{ml_oncol}. Thus, deep learning approaches should minimize the number of transcriptomic features selected to maximize both the accuracy and interpretability of such predictive biomarkers.

@@

% Materials and Methods
\section{Methods}
We developed a deep learning model to accurately classify cancer cell lines into therapeutic response groups using data from the Genomics of Drug Sensitivity in Cancer (GDSC) database \cite{gdsc}. Following data collection and curation, we utilized unsupervised clustering algorithms to define two groups based on chemotherapeutic response. Next, we employed a biologically agnostic feature selection algorithm, Boruta, to select statistically relevant genes for our neural network. These transcriptomic features were fed into a neural network that classifies patients into therapeutic response groups as defined by the clustering algorithms. The neural network architecture was optimized using a grid search and evaluated using 5-fold stratified cross-validation. See Fig. \ref{fig:pipeline} for an overview of the data analysis pipeline. The source code can be found on GitHub: \url{https://github.com/tig3r66/youreka_genes}.

% flowchart figure
\begin{figure}[!ht]
	\centering
	\begin{tikzpicture}[node distance=5.5em, auto]
	% blocks
	\node [block, text width=20em] (init) {Obtain data from the GDSC consortium (1,110 cell lines, 251 drugs).};
	\node [block, below of=init, text width=20em, yshift=-1em] (process) {Filter data: retain drugs with $>$80\% data for cell lines (548 cell lines, 117 drugs). Matrix normalization on IC$_{50}$ via z-score.};
	\node [block, below of=process, text width=20em, yshift=-1.5em] (kmeans) {Unsupervised clustering (k-means and principal component analysis) into 2 chemotherapeutic response groups.};
	\node [block, below of=kmeans, text width=20em, yshift=-1.5em] (append) {Appended GDSC Affymetrix Human Genome U219 expression data to the IC$_{50}$ matrix.};
	\node [block, below of=append, text width=20em, yshift=-1.5em] (feat) {Feature selection for relevant genes in RNA microarray data using Boruta algorithm. 300 genes selected at $\alpha=0.05$.};
	\node [block, below of=feat, text width=20em, yshift=-3.25em] (neural) {Architecture selection for number of hidden layers in the neural network followed by hyperparameter tuning via grid search. Neural networks with 1, 5, 10, and 15 hidden layers evaluated via 5-fold stratified cross-validation.};
	\node [block, below of=neural, text width=20em, yshift=-2.75em] (annot) {KEGG pathway enrichment of predictive genes selected by the Boruta algorithm.};

@@

Following evaluation of existing classifiers, we attempted to create defined cell line clusters using observed chemotherapeutic response of the pan-cancer cell line samples. We developed a Euclidean distance matrix for the retained cell lines based on their pan-chemotherapy response. This matrix was used to identify the minimum number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. Following this, k-means clustering was utilized to assign cell line candidates to appropriate therapeutic response cohorts. Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the pheatmap package \cite{pheatmap} in R. Separation between clusters was also visualized using principal component analysis with the factoextra package \cite{factoextra} in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons.

@@

\subsection{Feature Selection with Boruta}
In order to develop a transcriptomic model predictive of therapeutic response clusters, we retrieved expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line. Here, minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package \cite{affy} in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model developed from this large feature space is likely to exhibit high multicollinearity and subsequently overfit, compromising the generalizability of the model. In addition, the large feature space limits the efficiency of various feature selection algorithms\footnote{For example, univariate selection or recursive feature elimination.} as well as training and optimizing neural networks. We addressed these issues with the Boruta feature selection algorithm using the BorutaPy package \cite{liu} in Python 3. Boruta is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}

    \begin{subfigure}[t]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{Figures/roc/roc_combined.png}
    \end{subfigure}

    \caption{(\textbf{A}) Confusion matrices for the 1, 5, 10, and 15 hidden layer neural network models. The models classify cell line microarray datasets into chemotherapy response cohorts. (\textbf{B}) ROC curves for 1, 5, 10, and 15 hidden layers neural network models with confidence bands of $\pm 1$ standard error of the mean (sem). Each model was subject to 5-fold stratified cross-validation, and the mean ROC curve across all trials was plotted.}
    \label{fig:nnet}
\end{figure*}

@@

The dataset was randomly segregated using the Pareto principle \cite{pareto}; we reserved 80\% of the data for training and the remaining 20\% for validation. Model selection was performed by hyperparameter tuning using a grid search followed by 5-fold stratified cross-validation (Table \ref{tab:params}). We performed a grid search with 3-fold cross-validation on the training data (80\% of the dataset; 432 training samples, 541 overall) to determine the parameters that minimize the binary cross-entropy loss function. GridSearchCV from the scikit-learn library \cite{scikit-learn} was used to iterate through combinations of epochs, batch size, neurons per hidden layer, L2 regularization penalty, optimizer, and kernel initializer to find the optimal model (Table \ref{tab:params}). To prevent class imbalance during training, we used the Synthetic Minority Oversampling Technique (SMOTE) from the imblearn package \cite{imblearn} for Python 3. Each model’s performance on the testing dataset was evaluated by the Area Under Curve (AUC) value of the receiver operating characteristic (ROC) curve.

@@

\subsection{A neural network with 10 hidden layers accurately classifies patients into responder and non-responder cohorts}
Unsupervised learning in the form of the k-means clustering of the cancer cell line transcriptomes indicated substantially different responses to chemotherapies. Using these distinct therapy response cohorts, we developed a deep learning binary classifier to predict drug response groups based on transcriptomic data. We analyzed four neural network architectures: 1, 5, 10, and 15 hidden layers (Fig. \ref{fig:neural_vis}). Hyperparameter optimization via grid search returned similar results for each model: 25 epochs, batch size of 16, 100 neurons per hidden layer, L2 kernel regularization with $\lambda=10^{-3}$, stochastic gradient descent as the optimizer, and a normal kernel initializer. Following this, we validated the various architectures with 5-fold stratified cross-validation.

The neural network with 1 hidden layer performed poorly with 71.6\% accuracy. However, the 1 hidden layer model had the highest precision 95.5\% of the models (92.6\% for 5 hidden layers, 87.7\% for 10 hidden layers, 87.7\% for 15 hidden layers). The poor accuracy for the 1 hidden layer model was accounted for with a low negative predictive value (55.4\%; 80.5\% for 5 hidden layers, 100\% for 10 hidden layers, 100\% for 15 hidden layers). It was determined that the most accurate model was the neural network with 10 hidden layers (90.8\% accuracy; 71.6\% for 1 hidden layer, 88.1\% for 5 hidden layers, 85.3\% for 15 hidden layers; Fig. \ref{fig:nnet}).

An ROC curve was plotted for each of the neural network variants as an alternative evaluative method (Fig. \ref{fig:nnet}). The relative ratio between the model’s false positive rate and its true positive rate were averaged between 5 stratified k-folds. The mean AUC for the 5 trials was used to compare the four network architectures. However, the differences between the various architectures for AUC was not significantly different. Consequently, confusion matrices (Fig. \ref{fig:nnet}) and associated misclassification rates were used to pick the optimal model. The neural net with 10 hidden layers demonstrated the best performance overall with a 90.8\% accuracy.

@@

Following the determination of two defined chemotherapeutic groups based on the GDSC dataset, we used a biologically agnostic feature selection algorithm, Boruta, to reduce the original set of 16,382 genes to a subset of 300 genes and to limiting preliminary bias due to multicollinearity in the model. The genes were then fed into neural networks, which were optimized using a grid search (Table \ref{tab:params}). Each network was carefully monitored for overfitting during training (loss value observed to decrease during training across all models). To minimize overfitting, we applied L2 kernel regularization ($\lambda=10^{-3}$), batch normalization, and dropout layers with a 0.3 dropout rate to each hidden layer. To account for imbalanced classes during the cross-fold evaluations, we utilized SMOTE and stratified k-folds. We determined from confusion matrices and ROC curve analysis that the best network architecture utilized 10 hidden layers, and demonstrated a 90.8\% accuracy in classifying response groups (Fig. \ref{fig:nnet}). This validates our postulate that gene expression is a prime determinant of chemotherapeutic response, and that cell lines of similar gene expression profile respond similarly to most chemotherapies.

The comparatively lower accuracy of the neural network with 1 hidden layer (71.8\%) suggests that the therapeutic response cohorts cannot be separated by a linear classifier and that classical machine learning techniques are insufficient to capture the complexity of the dataset. Furthermore, the decrease in accuracy from 10 hidden layers (90.8\%) to 15 hidden layers (85.3\%) indicates overfitting. To this end, a 10 hidden layer model is the ideal depth to analyze the GDSC data. Interestingly, there is no significant difference between the mean AUC for these models (Fig. \ref{fig:nnet}). Given this finding, the confusion matrices were used to select the 10 hidden layer deep learning architecture as our model of choice.

@@

Future investigations will look to validate the predictive ability of the model to categorize chemotherapeutic response in various cancer types using the current transcriptomic signature. It is likely that the model accuracy will vary between therapy targets, and as such, further studies can make use of our project pipeline to create a stratified model whereby the drug class and target are additional inputs. In addition, biochemical features of gene targets may provide additional insights, especially for interpretability of the neural network output. It may also be interesting to select relevant genes using a different feature selector given that Boruta specifically operates on patterns of statistical relationships rather than biological relationships. Our use of Boruta was motivated by its efficacy demonstrated by prior studies of the algorithm as compared to other feature selectors \cite{boruta, deep_cell}. It is possible that a feature selection method informed by gene function and linkage disequilibrium could yield a different set of relevant genes.

% Conclusions
\section{Conclusions}
Using transcriptomic data from pan-cancer cell lines, two chemotherapeutic response clusters were identified via unsupervised learning in the form of k-means clustering. A feature selection algorithm was used to select a 300 gene subset which served as inputs to multiple neural networks. We determined that the network with 10 hidden layers was the most accurate model, producing a binary classifier to predict cell line therapy response with 90.8\% accuracy. Future studies will investigate the efficacy of our model to predict chemotherapy response in various cancer types and treatment contexts.

% Acknowledgements
\section*{Acknowledgements}
We wish to acknowledge the STEM Fellowship for organizing the 2020 Big Data Challenge, as well as Roche, SAS, Canadian Science Publishing, Digital Science, Altmetric, and Overleaf for their contributions that enabled this competition. We would like to thank our mentor, Dr. Daiva Nielsen for her feedback on our paper. We would also like to acknowledge Matthew Pietrosanu for his inputs.

@@

        \foreach \m/\l [count=\y] in {1,missing,100}
            \node [every neuron/.try, neuron \m/.try] (hidden-\m) at (1.5,-1.75-\y) {\foo{\l}};

        \foreach \m/\l [count=\y] in {1,missing,100}
            \node [every neuron/.try, neuron \m/.try] (hidden1-\m) at (3,-1.75-\y) {\hehe{\l}};

        \foreach \m/\l [count=\y] in {1,missing,100}
            \node [every neuron/.try, neuron \m/.try] (hidden2-\m) at (4.5,-1.75-\y) {\hehetwo{\l}};

        \foreach \m/\l [count=\y] in {1,missing,100}
            \node [every neuron/.try, neuron \m/.try] (hidden3-\m) at (6,-1.75-\y) {\hehethree{\l}};

        \foreach \m/\l [count=\y] in {1,missing,100}
			\node [every neuron/.try, neuron \m/.try] (hidden4-\m) at (7.5,-1.75-\y) {\hehefour{\l}};

@@

    	% input to hidden 1
        \foreach \l [count=\i] in {1,2,3,4,5,300}
            \foreach \k [count=\j] in {1,100}
              \draw [->] (input-\l) -- (hidden-\k);

    	% hidden 1 to hidden 2
        \foreach \l [count=\i] in {1,100}
            \foreach \k [count=\j] in {1,100}
              \draw [->] (hidden-\l) -- (hidden1-\k);

    	% hidden 2 to hidden 3
        \foreach \l [count=\i] in {1,100}
            \foreach \k [count=\j] in {1,100}
              \draw [->] (hidden1-\l) -- (hidden2-\k);

    	% hidden 3 to hidden 4
        \foreach \l [count=\i] in {1,100}
            \foreach \k [count=\j] in {1,100}
              \draw [->] (hidden2-\l) -- (hidden3-\k);

    	% hidden 4 to hidden 5
		\foreach \l [count=\i] in {1,100}
		\foreach \k [count=\j] in {1,100}
		\draw [->] (hidden3-\l) -- (hidden4-\k);

    	% hidden 5 to output
        \foreach \l [count=\i] in {1,100}
            \foreach \k [count=\j] in {1}
              \draw [->] (hidden4-\l) -- (output-\k);

@@

    \end{tikzpicture}
    \caption{Neural network architecture representation with 5 hidden layers (300 inputs). Inputs include the feature-selected genes from the cell line microarray dataset. Each hidden layer has an L2 kernel regularization ($\lambda=10^{-3}$) and a dropout rate of 0.3. All hidden layers are subject to batch normalization.}
    \label{fig:neural_vis}
\end{figure*}

>>>>

% Useful packages
\usepackage{amsmath, graphicx, natbib, multicol, booktabs, float, abstract, placeins, mathtools}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\bibliographystyle{unsrt}

@@

\begin{document}
% Abstract
\twocolumn[
	\maketitle
	\begin{abstract}
		Precision oncology approaches have recently begun to inform clinical decision-making in the context of chemotherapeutic regimen. High-throughput technology has enabled machine learning algorithms to utilize molecular characteristics of tumours to generate personalized treatments, revealing relationships between genomic alterations and drug sensitivity. However, precision oncology studies have yet to generate a predictive biomarker incorporating gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. Here we show that a 10 hidden layer neural network accurately classifies pan-cancer cell lines into responder and non-responder groups with respect to a pan-drug dataset with 89.0\% accuracy (AUC = 0.904, weighted F1 score = 0.89). From unsupervised clustering algorithms, we found that a cohort of cell line gene expression data from the Genomics of Drug Sensitivity in Cancer could be clustered into two response groups with significant difference in pan-drug chemotherapeutic sensitivity. We used the Boruta feature selection algorithm was used to identify genes of the highest predictive value; the 300 selected genes were enriched primarily for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using the selected genes, a deep learning neural network was developed to predict response groups. The high accuracy of the neural network validates our postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it suggests the potential utility of similar combinatorial biomarkers for the selection of potent candidate drugs. Given that not all tumours present with targetable features and are treated with less targeted and consequently more cytotoxic chemotherapies, our computational framework allows for the identification of candidate drugs for treatments that maximize therapeutic response and minimize cytotoxic burden. \vspace{1em}

@@

While previous studies have developed predictive biomarkers for specific drugs, the utility of these biomarkers is limited to particular patients and clinical contexts \cite{drug_sense}. That is, these studies are limited in terms of clinical generalizability to different chemotherapeutic regimens. However, a pan-drug predictive biomarker may provide significant clinical utility in the selection of candidate therapies for particular patients. Such a biomarker could be developed if tumours with similar expression have similar drug responses across most chemotherapies. The availability of pan-cancer cell line databases with \textit{in vitro} drug sensitivity analyses along with accompanying gene expression profiling provides an ideal model for such analyses \cite{gdsc}. However, the majority of previous drug sensitivity predictive biomarkers built on cell line databases have utilized classical machine learning techniques (ex: k-means clustering, regression), which fail to capture the dimensionality of available transcriptomic data. Furthermore, advanced deep learning algorithmic approaches that are capable of handling such dimensionality often fail to allow interpretability and consequently require transcriptomic data that is clinically infeasible \cite{ml_oncol}. Thus, deep learning approaches should minimize the number of selected transcriptomic features to maximize both the accuracy and interpretability of such predictive biomarkers. This would allow for a more robust pathway functional enrichment, and such analyses would provide a deeper understanding of the relative contributions of the selected transcriptomic features to the deep learning model.

@@

% Materials and Methods
\section{Methods}
We developed a deep learning model to accurately classify cancer cell lines into therapeutic response groups using data from the Genomics of Drug Sensitivity in Cancer (GDSC) database \cite{gdsc}. Following data collection and curation, we utilized unsupervised clustering algorithms to define two groups based on chemotherapeutic response. Next, we employed a biologically agnostic feature selection algorithm, Boruta, to select statistically relevant genes for our neural network. These transcriptomic features were fed into a neural network that classifies patients into therapeutic response groups as defined by the clustering algorithms. The neural network architecture was optimized using a grid search and evaluated using 5-fold stratified cross-validation. See Fig. \ref{fig:pipeline} for an overview of the data analysis pipeline. The source code can be found on GitHub: \url{https://github.com/tig3r66/youreka_genes}. \vspace{1em}

% flowchart figure
\begin{figure}[!ht]
	\centering
	\begin{tikzpicture}[node distance=6em, auto]
	% blocks
	\node [block, text width=20em] (init) {Obtain data from the GDSC consortium (1,110 cell lines, 251 drugs).};
	\node [block, below of=init, text width=20em, yshift=-1em] (process) {Filter data: retain drugs with $>$80\% data for cell lines (548 cell lines, 117 drugs). Matrix normalization on IC$_{50}$ via z-score.};
	\node [block, below of=process, text width=20em, yshift=-1.5em] (kmeans) {Unsupervised clustering (k-means and principal component analysis) into 2 chemotherapeutic response groups.};
	\node [block, below of=kmeans, text width=20em, yshift=-1.5em] (append) {Appended GDSC Affymetrix Human Genome U219 expression data to the IC$_{50}$ matrix.};
	\node [block, below of=append, text width=20em, yshift=-1.5em] (feat) {Feature selection for relevant genes in RNA microarray data using Boruta algorithm. 300 genes selected at $\alpha=0.05$.};
	\node [block, below of=feat, text width=20em, yshift=-3.25em] (neural) {Architecture selection for number of hidden layers in the neural network followed by hyperparameter tuning via grid search. Neural networks with 1, 5, 10, and 15 hidden layers evaluated via 5-fold stratified cross-validation.};
	\node [block, below of=neural, text width=20em, yshift=-2.75em] (annot) {KEGG pathway enrichment of predictive genes selected by the Boruta algorithm.};

@@

Following evaluation of existing classifiers, we attempted to create defined cell line clusters using observed chemotherapeutic response of the pan-cancer cell line samples. We developed a standardized Euclidean distance matrix for the retained cell lines based on their pan-chemotherapy response. This matrix was used to identify the minimum number of clusters (the elbow method was used as an heuristic for determining the number of clusters) capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. Following this, k-means clustering was utilized to assign cell line candidates to appropriate therapeutic response cohorts. Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the pheatmap package \cite{pheatmap} in R. The heatmap displayed 2 groups clustered using k-means clustering. Separation between clusters was also visualized using principal component analysis with the factoextra package \cite{factoextra} in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons.

@@

\subsection{Feature Selection with Boruta}
In order to develop a transcriptomic model predictive of therapeutic response clusters, we retrieved expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line. Here, minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package \cite{affy} in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model developed from this large feature space is likely to exhibit high multicollinearity and subsequently overfit, compromising the generalizability of the model. In addition, the large feature space limits the efficiency of various feature selection algorithms (for example, univariate selection or recursive feature elimination) as well as training and optimizing neural networks. We addressed these issues with the Boruta feature selection algorithm using the BorutaPy package \cite{liu} in Python 3. Boruta is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}

% neural net evaluation figure
\begin{figure*}[!ht]
    \centering
    \begin{subfigure}[t]{\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/roc/final_roc.png}
	\end{subfigure}

    \begin{subfigure}[t]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{Figures/confusion_matrix/fcm_combined.png}
    \end{subfigure}

    \caption{(\textbf{A}) ROC curves for 1, 5, 10, and 15 hidden layers neural network models. Each model was subject to 5-fold stratified cross-validation on the training set, and the mean ROC curve for the testing set across all trials was plotted. (\textbf{B}) Confusion matrices for the 1, 5, 10, and 15 hidden layer neural network models on the testing set. The threshold for classification determiend using Youden's index (thresholds were 0.17, 0.97, 0.49, and 0.45 for neural networks with 1, 5, 10, and 15 hidden layers respectively). The models classify cell line microarray datasets into chemotherapy response cohorts.}
    \label{fig:nnet}
\end{figure*}

@@

The dataset was randomly segregated using the Pareto principle \cite{pareto}; we reserved 80\% of the data for training and the remaining 20\% for testing. Model selection was performed by hyperparameter tuning using a grid search followed by 5-fold stratified cross-validation on the training data (Table \ref{tab:params}). We performed a grid search with 3-fold cross-validation on the training data (80\% of the dataset; 432 training samples, 541 overall) to determine the parameters that minimize the binary cross-entropy loss function. GridSearchCV from the scikit-learn library \cite{scikit-learn} was used to iterate through combinations of epochs, batch size, neurons per hidden layer, L2 regularization penalty, optimizer, and kernel initializer to find the optimal model (Table \ref{tab:params}). To prevent class imbalance during training, we used the Synthetic Minority Oversampling Technique (SMOTE) from the imblearn package \cite{imblearn} for Python 3. Each model’s performance on the testing dataset was evaluated by the Area Under Curve (AUC) value of the receiver operating characteristic (ROC) curve.

Following grid search and cross-validation, we evaluated the performance of our model on the testing data. Youden's index from the ROC curves corresponding to each neural network was used to determine the optimal threshold to classify cell lines into chemotherapeutic response groups. The formula for Youden's index is defined as $$J \doteq \text{sensitivity} - \text{specificity} - 1$$ All indices were obtained from the ROC curves by an iterative search for the maximum $J$. That is, $$J = \text{argmax(TPR} - \text{FPR)}$$ where TPR and FPR are arrays containing the true positive and false positive fate corresponding to each point on the ROC curve. The threshold corresponding to $J$ was used to classify cell lines.

@@

\subsection{A neural network with 10 hidden layers accurately classifies patients into responder and non-responder cohorts}
Unsupervised learning via k-means clustering of the cancer cell line transcriptomes indicated substantially different responses to chemotherapies. Since centroids are initially randomly placed, we ran the algorithm several times, and each iteration returned similar results. Using these distinct therapy response cohorts, we developed a deep learning binary classifier to predict drug response groups based on transcriptomic data. We analyzed four neural network architectures: 1, 5, 10, and 15 hidden layers (Fig. \ref{fig:neural_vis}). Hyperparameter optimization via grid search returned similar results for each model: 50 epochs, batch size of 32, 50 neurons per hidden layer, L2 kernel regularization with $\lambda=10^{-3}$, Adam as the optimizer, and a normal kernel initializer. Following this, we validated the various architectures with 5-fold stratified cross-validation.

On the training data, the neural network with 5 hidden layers performed the best on the validation fold with 93.1\% accuracy (88.8\% for 1 hidden layer, 84.5\% for 10 hidden layers, 56.9\% for 15 hidden layers). The mean AUC across the 5 folds were $0.965 \pm 0.017$, $0.957 \pm 0.0144$, $0.913 \pm 0.026$, and $0.515 \pm 0.115$ for the neural networks with 1, 5, 10, and 15 hidden layers respectively.

An ROC curve was plotted for each of the neural network, and the threshold which maximizes Youden's index was determined (Fig. \ref{fig:nnet}). Here the relative ratio between the model’s false positive rate and its true positive rate were averaged between 5 stratified k-folds. The mean AUC for the 5 trials was used to compare the four network architectures. While there were no significant differences between AUC for the 1, 5, and 10 hidden layer models, the model with 15 hidden layers performed significantly worse compared to the rest.

The results were different for the testing data. The neural network with 10 hidden layers performed the best with an accuracy of 89.0\% (82.6\% for 1 hidden layer, 73.4\% for 5 hidden layers, and 57.8\% for 15 hidden layers). The AUC was 0.885, 0.884, 0.904, and 0.583 for the neural networks with 1, 5, 10, and 15 hidden layers respectively. Of note, the neural network with 15 hidden layers classified all cell lines into the responder group. The weighted F1 scores for the neural networks with 1, 5, 10, and 15 layers were 0.83, 0.73, 0.89, and 0.42 respectively.

@@

Following the determination of two defined chemotherapeutic groups based on the GDSC dataset, we used a biologically agnostic feature selection algorithm, Boruta, to reduce the original set of 16,382 genes to a subset of 300 genes and to limiting preliminary bias due to multicollinearity in the neural network models. The genes were then fed into neural networks, which were optimized using a grid search (Table \ref{tab:params}). Each network was carefully monitored for overfitting during training (loss value on the validation set was observed to decrease during training across all models). To minimize overfitting, we applied L2 kernel regularization ($\lambda=10^{-3}$), batch normalization, and dropout layers with a 0.3 dropout rate to each hidden layer. To account for imbalanced classes during the cross-fold evaluations, we utilized SMOTE and stratified k-folds. We determined from confusion matrices and ROC curve analysis that the best network architecture utilized 10 hidden layers, and demonstrated a 89.0\% accuracy in classifying response groups (Fig. \ref{fig:nnet}). This validates our postulate that gene expression is a prime determinant of chemotherapeutic response, and that cell lines of similar gene expression profile respond similarly to most chemotherapies.

\Large{\textbf{FIX PARA BELOW}}\\
\normalsize
The comparatively lower accuracy of the neural network with 1 hidden layer (71.8\%) suggests that the therapeutic response cohorts cannot be separated by a linear classifier and that classical machine learning techniques are insufficient to capture the complexity of the dataset. Furthermore, the decrease in accuracy from 10 hidden layers (90.8\%) to 15 hidden layers (85.3\%) indicates overfitting. To this end, a 10 hidden layer model is the ideal depth to analyze the GDSC data. Interestingly, there is no significant difference between the mean AUC for these models (Fig. \ref{fig:nnet}). Given this finding, the confusion matrices were used to select the 10 hidden layer deep learning architecture as our model of choice.

@@

Future investigations will look to validate the predictive ability of the model to categorize chemotherapeutic response in various cancer types using the current transcriptomic signature. It is likely that the model accuracy will vary between therapy targets, and as such, further studies can make use of our project pipeline to create a stratified model whereby the drug class and target are additional inputs. In addition, biochemical features of gene targets may provide additional insights, especially for interpretability of the neural network output. It may also be interesting to select relevant genes using a different feature selector given that Boruta specifically operates on patterns of statistical relationships rather than biological relationships. As such, Boruta is sensitive to hyperparameter values. Our use of Boruta was primarily motivated by its efficacy demonstrated by prior studies of the algorithm as compared to other feature selectors \cite{boruta, deep_cell}. It is possible that a feature selection method informed by gene function and linkage disequilibrium could yield a different set of relevant genes.

% Conclusions
\section{Conclusions}
Using transcriptomic data from pan-cancer cell lines, two chemotherapeutic response clusters were identified via unsupervised learning in the form of k-means clustering. The Boruta feature selection algorithm was used to select a 300 gene subset which served as inputs to multiple neural networks. We determined that a network with 10 hidden layers was the most accurate model, producing a binary classifier to predict cell line therapy response with 89.0\% accuracy. Future studies will investigate the efficacy of our model to predict chemotherapy response in various cancer types and treatment contexts.

% Acknowledgements
\section*{Acknowledgements}
We wish to acknowledge the STEM Fellowship for organizing the 2020 Big Data Challenge, as well as Roche, SAS, Canadian Science Publishing, Digital Science, Altmetric, and Overleaf for their contributions that enabled this competition. We would like to thank our mentor, Dr. Daiva Nielsen for her feedback on our paper. We would also like to acknowledge Matthew Pietrosanu, Boya Peng, Danyi Liu, Stephen Styles, and Ree Yang for their inputs.

@@

        \foreach \m/\l [count=\y] in {1,missing,50}
            \node [every neuron/.try, neuron \m/.try] (hidden-\m) at (1.5,-1.75-\y) {\foo{\l}};

        \foreach \m/\l [count=\y] in {1,missing,50}
            \node [every neuron/.try, neuron \m/.try] (hidden1-\m) at (3,-1.75-\y) {\hehe{\l}};

        \foreach \m/\l [count=\y] in {1,missing,50}
            \node [every neuron/.try, neuron \m/.try] (hidden2-\m) at (4.5,-1.75-\y) {\hehetwo{\l}};

        \foreach \m/\l [count=\y] in {1,missing,50}
            \node [every neuron/.try, neuron \m/.try] (hidden3-\m) at (6,-1.75-\y) {\hehethree{\l}};

        \foreach \m/\l [count=\y] in {1,missing,50}
			\node [every neuron/.try, neuron \m/.try] (hidden4-\m) at (7.5,-1.75-\y) {\hehefour{\l}};

@@

    	% input to hidden 1
        \foreach \l [count=\i] in {1,2,3,4,5,300}
            \foreach \k [count=\j] in {1,50}
              \draw [->] (input-\l) -- (hidden-\k);

    	% hidden 1 to hidden 2
        \foreach \l [count=\i] in {1,50}
            \foreach \k [count=\j] in {1,50}
              \draw [->] (hidden-\l) -- (hidden1-\k);

    	% hidden 2 to hidden 3
        \foreach \l [count=\i] in {1,50}
            \foreach \k [count=\j] in {1,50}
              \draw [->] (hidden1-\l) -- (hidden2-\k);

    	% hidden 3 to hidden 4
        \foreach \l [count=\i] in {1,50}
            \foreach \k [count=\j] in {1,50}
              \draw [->] (hidden2-\l) -- (hidden3-\k);

    	% hidden 4 to hidden 5
		\foreach \l [count=\i] in {1,50}
		\foreach \k [count=\j] in {1,50}
		\draw [->] (hidden3-\l) -- (hidden4-\k);

    	% hidden 5 to output
        \foreach \l [count=\i] in {1,50}
            \foreach \k [count=\j] in {1}
              \draw [->] (hidden4-\l) -- (output-\k);

@@

    \end{tikzpicture}
    \caption{Neural network architecture representation with 5 hidden layers (300 inputs). Inputs include the feature-selected genes from the cell line microarray dataset. Each hidden layer is batch-normalized with an L2 kernel regularization penalty ($\lambda=10^{-3}$) and a dropout rate of 0.3.}
    \label{fig:neural_vis}
\end{figure*}

	\begin{subfigure}[t]{\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/roc/cv_roc.png}
	\end{subfigure}

	\caption{(\textbf{A}) Confusion matrices for the 1, 5, 10, and 15 hidden layer neural network models on the validation set. The models classify cell line microarray datasets into chemotherapy response cohorts. (\textbf{B}) ROC curves for 1, 5, 10, and 15 hidden layers neural network models with confidence bands of $\pm 1$ standard error of the mean (sem). Each model was subject to 5-fold stratified cross-validation on the training set, and the mean ROC curve for the validation set across all trials was plotted.}
	\label{fig:nnet_train}
\end{figure*}
